---
title: "Founding principles"
vignette: >
  %\VignetteIndexEntry{Founding principles}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
# library(cgrasch)
```

## Introduction

-   Imagine a high jumper who has a PB of 2 metres.

-   The chance, or the odds of success can be expressed as the jumper's ability minus the difficulty of the task.

-   Odds is expressed as the probability of success over the probability of failure.

-   $odds = \frac{P_1}{P_0}$

## Some algebra and why

-   Odds is the probability of success over the probability of failure $odds = \frac{P_1}{P_0}$

-   Taking the natural log makes odds into log odds

-   It maps the probabilities to the whole number line -$\infty$ to $\infty$ and makes the log odds of a 50:50 chance zero

-   From $log\ odds = ln(\frac{P}{1 - P})$ ...

-   We get $\beta - \delta = ln(\frac{P}{1 - P})$ or $ln(\frac{P_1}{P_0})$

## Steps for whom at school girls > maths

$$ 
\begin{aligned}
\beta - \delta &= \frac{P_1}{P_0}
\\&= ln(\frac{P}{1-P})
\\e^{(\beta - \delta)} &= \frac{P}{1 - P}
\\P &= e^{\beta - \delta}(1 - P)
\\&= e^{\beta - \delta} -Pe^{\beta - \delta}
\\e^{\beta - \delta} &= P + Pe^{\beta - \delta}
\\&= P(1 + e^{\beta - \delta})
\\P &= \frac{e^{\beta - \delta}}{1 + e^{\beta - \delta}}
\end{aligned}
$$

-   This means that if your ability matches the difficulty, i.e. $\beta - \delta = 0$, the probability of success is .5

-  I.e. $e(0) = 1$, so $1 / 1 + 1 = 0.5$

## Example for a single item

```{r}
#| echo: true

b <- 7
tibble::tibble(d = seq(0, 14, .1)) |>
  dplyr::mutate(p1 = exp(b - d) / (1 + exp(b - d))) |>
  ggplot2::ggplot(ggplot2::aes(y = d, x = p1)) +
  ggplot2::geom_point() +
  ggplot2::geom_hline(yintercept = b, linetype = "dashed") +
  ggplot2::theme_minimal() +
  ggplot2::labs(y = "difficulty", x = "probability of success")

```
