---
title: "How Rasch measures are calculated using JMLE"
vignette: >
  %\VignetteIndexEntry{How Rasch measures are calculated using JMLE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

This is an r implementation of an Excel demo by Mark Moulton, which can be found at https://www.rasch.org/moulton.htm.

```{r}
#| label: setup
library(cgrasch)
library(tidyverse)
library(flextable)
```

```{r}
#| label: data

# Use package data of a simple set of 10 candidates' responses to dichotomous
# items. 
candidate_data <- dichotomous

```

Persons A to I are assigned a probability of success between .1 and .9 and their performance simulated on 10 dichotomous items. The modelling can't handle perfect or zero scores. Candidate A is given an NA for item 1 to simulate missing data. [This ability to handle missing data is the foundation of the concept of linking.]

```{r}
#| echo: false

candidate_data |> 
  pivot_wider(names_from = item, values_from = response) |> 
  flextable()

```

First, calculate the proportion correct (P) for each candidate and convert to person ability logits by taking the natural log of the odds of success.

$$ person\ ability\ logit = ln(\frac{p}{1-p}) $$

Do the same for items, calculating proportion *incorrect* and converting to item difficulty logits. Also, calculate the adjusted item difficulty logits so that they have a mean of 0.00 logits. This anchors the scale on zero and makes it possible to compute expected values that approximate to observed values.

$$ item\ difficulty\ logit = ln(\frac{1 - p}{p}) $$

```{r}
#| label: summary

summary_candidates <- candidate_data |> 
  group_by(person) |> 
  reframe(P = sum(response, na.rm = TRUE)/n()) |> 
  mutate(ability_logit = log(P/(1-P)))

summary_items <- candidate_data |> 
  group_by(item) |> 
  reframe(P = sum(response, na.rm = TRUE)/n()) |> 
  mutate(
    difficulty_logit = log((1-P)/P),
    adj_difficulty_logit = difficulty_logit - 
      mean(difficulty_logit, na.rm = TRUE)
  )

```

```{r}
#| label: summary-candidates
#| echo: false

summary_candidates |> 
  mutate(across(where(is.numeric), ~ round(., 2))) |> 
  flextable()

```

\

```{r}
#| label: summary-items
#| echo: false

summary_items |> 
  mutate(across(where(is.numeric), ~ round(., 2))) |> 
  flextable()

```

\

The next step is to calculate the expected values for each person and item. This value is calculated as the sum of each possible value times its probability. For person A on item 1, the probability is given as:

$$ P_{A1=1} = \frac{e^{(person\ logit - item\ logit)}}{1 + e^{(person\ logit - item\ logit)}} $$

```{r}
#| label: iteration-one

iteration_1 <- summary_candidates |>
    select(person, ability_logit) |>
    mutate(
        data = list(
            summary_items |>
                select(item, adj_difficulty_logit)
        )
    ) |>
    unnest(data) |>
    mutate(
        expected_value = exp(ability_logit - adj_difficulty_logit) /
            (1 + exp(ability_logit - adj_difficulty_logit)),
        variance = expected_value * (1 - expected_value)
    )

```

```{r}
#| label: iteration-one-table
#| echo: false

iteration_1 |> 
  select(person, item, expected_value) |> 
  pivot_wider(names_from = item, values_from = expected_value) |> 
  mutate(across(where(is.numeric), ~ round(., 2))) |> 
  flextable()

```

Next we calculate the variance of each estimate, which is the standard error squared.

$$ V = p(1 - p) $$

```{r}
#| label: variance-table

iteration_1 |> 
  select(person, item, variance) |> 
  pivot_wider(names_from = item, values_from = variance) |> 
  mutate(across(where(is.numeric), ~ round(., 2))) |> 
  flextable()

```

Then calculate the sum of variances by candidate and multiply by -1. This figure will be used to adjust the ability logit later in this first iteration. Also, calculate the sum of variances for each item and multiply -1. These figures are second derivatives of the likelihood function; they are used as the denominator when tweaking the difficulty logit to maximise fit.

```{r}
#| label: sum-variances

# I've done this differently below and not used these values.

sum_person_variances <- iteration_1 |> 
  dplyr::group_by(person) |> 
  dplyr::reframe(sum_variances= sum(variance) * -1)
  
sum_item_variances <- iteration_1 |> 
  group_by(item) |> 
  reframe(sum_variances= sum(variance) * -1)

```

Next, calculate the residuals between the iterations.

```{r}
#| label: residuals-1

residuals_1 <- iteration_1|> 
  left_join(
    candidate_data,
    by = join_by(person, item)
  ) |> 
  mutate(residual = response - expected_value)

```

```{r}
#| label: residuals-table
#| echo: false

residuals_1 |> 
  select(person, item, residual) |> 
  pivot_wider(names_from = item, values_from = residual) |> 
  mutate(across(where(is.numeric), ~ round(., 2))) |> 
  flextable()

```

Now, compute the sum of the residuals by candidate. The goal is to adjust the ability and difficulty logit measures until this sum of residuals equals zero for each candidate. This solution is known as the *Maximum Likelihood* solution. It is nearly the same as the *Least Squares* solution.

```{r}
#| label: maximum-likelihood

residuals_1 |> 
  group_by(person) |> 
  reframe(sum_residuals = sum(residual, na.rm = TRUE)) |> 
  reframe(ss = sum(sum_residuals^2))
  
```

Use the sum of the residuals and the sum of the variances calculated to adjust the ability scores.

$$ new\ ability\ logit = ability\ logit - \left(\frac{sum\ residuals}{sum\ variances}\right) $$

```{r}
#| label: adjust-ability

adjusted_ability_1 <- summary_candidates |> 
  dplyr::left_join(
    residuals_1 |> 
      dplyr::group_by(person) |> 
      dplyr::reframe(sum_residuals = sum(residual, na.rm = TRUE)),
    by = dplyr::join_by(person)
  ) |> 
  dplyr::left_join(
    residuals_1 |> 
      dplyr::group_by(person) |> 
      dplyr::reframe(sum_variances = sum(variance)),
    by = dplyr::join_by(person)
  ) |> 
  dplyr::mutate(
    new_adj_ability_logit = ability_logit - (sum_residuals / sum_variances)
  )

adjusted_ability_1 |> 
  flextable::flextable()

```

As the sum of residuals (divided by the sum of variances which standardizes it) approaches zero, the amount by which you need to adjust the previous ability estimate also approaches zero, and you converge on the maximum likelihood solution.

Do the same for the item difficulties, and adjust them so that their mean is zero.

```{r}
#| label: adjust-difficulty

adjusted_difficulty_1 <- summary_items |> 
  dplyr::left_join(
    residuals_1 |> 
      dplyr::group_by(item) |> 
      dplyr::reframe(sum_residuals = sum(residual, na.rm = TRUE)),
    by = dplyr::join_by(item)
  ) |> 
  dplyr::left_join(
    residuals_1 |> 
      dplyr::group_by(item) |> 
      dplyr::reframe(sum_variances = sum(variance)),
    by = dplyr::join_by(item)
  ) |> 
  dplyr::mutate(
    new_difficulty_logit = adj_difficulty_logit  - (sum_residuals / sum_variances),
    new_adj_difficulty_logit = new_difficulty_logit - 
      mean(new_difficulty_logit)
  )

adjusted_difficulty_1 |> 
  flextable::flextable()

```

Now, keep iterating this process by  Computing a new round of expected values, but this time use the improved set of person ability logits and item difficulty logits. Keep iterating until the sum of the square residuals is zero or close to it. The function `cgrasch::cg_rm()` completes as many iterations as is required to reach a specified sensitivity. The default is set to .00001. The function provides all iterations of the ability and difficult estimates, the final sum of square residuals value and separate tables of the final ability and difficulty estimates.

```{r}
#| label: cgrasch-model

model <- dichotomous |> 
  cg_rm()

```

\

```{r}
#| label: abilities

model$final_person_abilities |> 
  flextable::flextable()

```

\

```{r}
#| label: difficulties

model$final_item_difficulties |> 
  flextable::flextable()

```

